Synthesis of production rules for glycans

### 2020-10-15 papers/sugar-synth/algo.tex
variations first cut

\begin{wrapfigure}{r}{0.37\textwidth}
    \vspace{-12mm}
        \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid,scale=0.5]
    \node[loc] (v0) {$v_0$};
    \node[sqloc, below right of=v0] (v1) {$v_e$};
    \draw [->] plot [smooth] coordinates { (v0.south)
          ($(v0.south)+(3mm,-2mm)$)
          ($(v0.south)+(10mm,1mm)$)
          % ($(v0.south)+(9mm,-6mm)$)
          % ($(v0.east)+(8mm,-8mm)$)
          (v1.north) };
    % \path[->] (v0) edge (v1);
    % \path[->,dashed] (v1) edge (v2);
    % \path[->] (v1) edge (v3);
    \fill[fill=gray,opacity=0.2] 
    (0,0) -- ++(-1.5cm,-1.7cm) -- ++(1.2cm,-1.2cm) -- ++(6cm,0cm) -- cycle;
    \fill[fill=gray,opacity=0.4] 
    (v1.center) -- ++(-1.5cm,-1.5cm) -- ++(3cm,0cm) -- cycle;
    \draw[dashed] (-1.5cm,-1.7cm) -- ++(-1.2cm,-1.2cm) -- ++(2.4cm,0cm) -- cycle;
  \end{tikzpicture}
  \caption{Parts of production rules in the rule templates.}
  \label{fig:rule-parts}
  \vspace{-8mm}
\end{wrapfigure}
A node in a production rule can be in four situations.
In Figure~\ref{fig:rule-parts}, we illustrate the situations.
The first situation is when a node position is in the expanding part.
The dark gray nodes are the expanding nodes.
The second situation is when a node position is not in the rule.
The dashed area are the absent nodes.
Among the matching nodes, we have two cases.
% , namely if the expanding
% nodes are the descendants of the matching node or not.
The third situation is when a node position is in the matching part
and has expanding decadents.
The nodes on the solid path from $v_0$ to the root $v_e$ of the expanding part are in the third situation.
% the nodes that are ancestors of expanding part among the matching nodes.
Finally, the fourth situation is the rest of node positions
in the matching part, which is the light gray area.
% The nodes in the rest of the light gray area are in the fourth situation.
A variable is mapped to a node to encode the four situations.
Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
indicate the set of four situations.
Let $KVars$ be the set of uncountable variables with domain $K$. 
% We will use the pool of $KVars$ to add variables to the templates.
The templates are defined as follows.

@@

%
Similar to the rule templates, not all assignments to molecule template variables
are valid.
We add the following conditions for
valid assignments for molecule template $\hat{m} = (V, \nu, C, \hat{v_0})$.
% The first  two conditions encode the tree structure of the molecule and
% matching the number of children with label arity for each node.
% The second encodes that the height of the molecule is large enough.
% The last encodes that new molecule must not be the same as any molecule in $\mu$.
\begin{enumerate}
\item If a node is present, then the parent of the nodes is also present.\\
  $
  \Land_{i\in[1,w]}\Land_{{\text{internal node } v \in V}} (\nu(C(v,i) ) \neq \bot \limplies \nu( v ) \neq \bot )  
  $
\item The children count of a node matches with the arity of the labeled sugar.\\
  $
  \Land_{s \in S} \Land_{i \in (arity(s),w] }\Land_{v \in V} (\nu(v) = s \limplies \kappa( C(v,i) ) = Absent )
  $
% \item At least one of the leaves is present. This is for
%   avoiding the trivial solution.\\
%   $
%   \Lor_{ v \in V \text{ and } depth(v) > d' } \nu( v ) \neq \bot,
%   $
%   where $d'$ is a number close to and smaller than $height(\hat{v_0})$.
%   This allows us to look for all molecules that are close
%   to height $h$.
%   We choose $d' = h-d$, where $d$ is the height of the rule templates.
  % This $d'$ ensures that we do not miss producible molecules that
  % can not have exact height $h$ but is slightly larger than $h$.
  % Then, there must be another producible
  % molecule that is a subtree of the molecule and has height in
  % between $h$ and $h-d$
  % because the maximum depth of a rule is $d$.
\item We find a molecule that is not in $\mu$.
  We encode $\Lor_{{m = (\_,\_,\_, v_0)\in \mu}}  Neq(\hat{v_0},v).$, where
   predicate $Neq$ be recursively defined as follows.\\
  $Neq(v, v') :=\; \nu(v) \neq M(v') \lor
  \Lor_{{i \in [1,w]}} Neq( C(v,i), C(v',i ) )$\\
  $Neq(v,\bot) :=\;  \nu(v) \neq \bot$
  % \begin{align*}
  % Neq(v, v') :=\;&   \nu(v) \neq M(v') \lor
  % \Lor_{\mathclap{i \in [1,w]}} Neq( C(v,i), C(v',i ) )\\
  %   Neq(v,\bot) :=\;&  \nu(v) \neq \bot
  % \end{align*}
  % The constraint is
  % % \begin{align*}
  % % \end{align*}
  % In our implementation, 
  % we have an option to also ignore the subtrees of
  % the molecules of $\mu$.
\end{enumerate}
%
In our method, the call to $\textsc{MolTemplateCorrectness}$ at
line~\ref{line:molCorr} implements above constraints and store them in
$mCons$.

@@

Once a rule is applied to a molecule, all the
nodes inside the expanding part are added together.
Therefore, there should be no cuts with in the set of added nodes.
Furthermore, if there are children nodes below the added nodes,
they must be added due to the application of some other rule.
Therefore, the children are at the cut points.
We call the above requirement {\em cut pattern}. 
$\textsc{MatchCuts}$ is a recursive function over the trees, and matches
cut patterns between the molecule node $v$ and the rule template node $v_r$.
The cuts must occur whenever nodes of the rule template transition
from present to absent.
For helping to detect the transition,
the third parameter is a constraint that encodes that the parent of $v_r$
is absent or not.
% If $v$ does not exist, then there is no constraint to add therefore
% the procedure returns $\ltrue$ at line~\ref{line:mcut-no-v}.
% If $v_r$ does not exist, then it returns constraints encoding
% that if the parent of $v_r$ was not absent, then $v$ must be at a cut at line~\ref{line:mcut-no-vr}.
% If both $v$ and $v_r$ exist, then we add constraints
% that if the parent of $v_r$ is not absent, $v_r$ is
% flagged to be absent in the learned rule
% if and only if $v$ is the cut points at line~\ref{line:mcut-cons}.
% $\textsc{MatchCuts}$ recursively traverses the children of $v$ and $v_r$
% and generates the constraints for the cut pattern.
% The constraints added by the procedure are guarded by the presence of parents
% because we need to enable constraints only when there is a possibility of transition
% from presence to absence.
The call to $\textsc{MatchCuts}$ in \textsc{EncodeP} passes $\lfalse$ as the third input.
Even if the parent of $v_r$ in \textsc{EncodeP} is present,
$v_r$ is a cut point adding cut pattern constraints for the node will create inconsistency.
Therefore, we are passing $\lfalse$.

>>>>

\begin{wrapfigure}{r}{0.37\textwidth}
    \vspace{-3mm}
        \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid,scale=0.5]
    \node[loc] (v0) {$v_0$};
    \node[sqloc, below right of=v0] (v1) {$v_e$};
    \draw [->] plot [smooth] coordinates { (v0.south)
          ($(v0.south)+(3mm,-2mm)$)
          ($(v0.south)+(10mm,1mm)$)
          % ($(v0.south)+(9mm,-6mm)$)
          % ($(v0.east)+(8mm,-8mm)$)
          (v1.north) };
    % \path[->] (v0) edge (v1);
    % \path[->,dashed] (v1) edge (v2);
    % \path[->] (v1) edge (v3);
    \fill[fill=gray,opacity=0.2] 
    (0,0) -- ++(-1.5cm,-1.7cm) -- ++(1.2cm,-1.2cm) -- ++(6cm,0cm) -- cycle;
    \fill[fill=gray,opacity=0.4] 
    (v1.center) -- ++(-1.5cm,-1.5cm) -- ++(3cm,0cm) -- cycle;
    \draw[dashed] (-1.5cm,-1.7cm) -- ++(-1.2cm,-1.2cm) -- ++(2.4cm,0cm) -- cycle;
  \end{tikzpicture}
  \caption{Parts of production rules in the rule templates.}
  \label{fig:rule-parts}
  \vspace{-8mm}
\end{wrapfigure}
A node in a production rule can be in four situations.
In Figure~\ref{fig:rule-parts}, we illustrate the situations.
The first situation is when a node position is in the expanding part.
The dark gray nodes are the expanding nodes.
The second situation is when a node position is not in the rule.
The dashed area are the absent nodes.
Among the matching nodes, we have two cases.
% , namely if the expanding
% nodes are the descendants of the matching node or not.
The third situation is when a node position is in the matching part
and has expanding decadents.
The nodes on the solid path from $v_0$ to the root $v_e$ of the expanding part are in the third situation.
% the nodes that are ancestors of expanding part among the matching nodes.
Finally, the fourth situation is the rest of node positions
in the matching part, which is the light gray area.
% The nodes in the rest of the light gray area are in the fourth situation.
A variable is mapped to a node to encode the four situations.
Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
indicate the set of four situations.
Let $KVars$ be the set of uncountable variables with domain $K$. 
% We will use the pool of $KVars$ to add variables to the templates.
The templates are defined as follows.

@@

%
Similar to the rule templates, not all assignments to molecule template variables
are valid.
We add the following conditions for
valid assignments for molecule template $\hat{m} = (V, \nu, C, \hat{v_0})$.
% The first  two conditions encode the tree structure of the molecule and
% matching the number of children with label arity for each node.
% The second encodes that the height of the molecule is large enough.
% The last encodes that new molecule must not be the same as any molecule in $\mu$.
\begin{enumerate}
\item If a node is present, then the parent of the nodes is also present.\\
  $
  \Land_{i\in[1,w]}\Land_{{\text{internal node } v \in V}} (\nu(C(v,i) ) \neq \bot \limplies \nu( v ) \neq \bot )  
  $
\item The children count of a node matches with the arity of the labeled sugar.\\
  $
  \Land_{s \in S} \Land_{i \in (arity(s),w] }\Land_{v \in V} (\nu(v) = s \limplies \kappa( C(v,i) ) = Absent )
  $
% \item At least one of the leaves is present. This is for
%   avoiding the trivial solution.\\
%   $
%   \Lor_{ v \in V \text{ and } depth(v) > d' } \nu( v ) \neq \bot,
%   $
%   where $d'$ is a number close to and smaller than $height(\hat{v_0})$.
%   This allows us to look for all molecules that are close
%   to height $h$.
%   We choose $d' = h-d$, where $d$ is the height of the rule templates.
  % This $d'$ ensures that we do not miss producible molecules that
  % can not have exact height $h$ but is slightly larger than $h$.
  % Then, there must be another producible
  % molecule that is a subtree of the molecule and has height in
  % between $h$ and $h-d$
  % because the maximum depth of a rule is $d$.
\item We find a molecule that is not in $\mu$.
  We encode $\Land_{{(\_,\_,\_, v_0)\in \mu}}  Neq(\hat{v_0},v_0)$, where
   predicate $Neq$ be recursively defined as follows.\\
  $Neq(v, v') :=\; \nu(v) \neq M(v') \lor
  \Lor_{{i \in [1,w]}} Neq( C(v,i), C(v',i ) )$\\
  $Neq(v,\bot) :=\;  \nu(v) \neq \bot$
  % \begin{align*}
  % Neq(v, v') :=\;&   \nu(v) \neq M(v') \lor
  % \Lor_{\mathclap{i \in [1,w]}} Neq( C(v,i), C(v',i ) )\\
  %   Neq(v,\bot) :=\;&  \nu(v) \neq \bot
  % \end{align*}
  % The constraint is
  % % \begin{align*}
  % % \end{align*}
  % In our implementation, 
  % we have an option to also ignore the subtrees of
  % the molecules of $\mu$.
\end{enumerate}
%
In our method, the call to $\textsc{MolTemplateCorrectness}$ at
line~\ref{line:molCorr} implements above constraints and store them in
$mCons$.

@@

Once a rule is applied to a molecule, all the
nodes inside the expanding part are added together.
Therefore, there should be no cuts with in the set of added nodes.
Furthermore, if there are children nodes below the added nodes,
they must be added due to the application of some other rule.
Therefore, the children are at the cut points.
We call the above requirement {\em cut pattern}. 
$\textsc{MatchCuts}$ is a recursive function over the trees, and matches
cut patterns between the molecule node $v$ and the rule template node $v_r$.
The cuts must occur whenever nodes of the rule template transition
from present to absent.
For helping to detect the transition,
the third parameter is a constraint that encodes that the parent of $v_r$
is absent or not.
% If $v$ does not exist, then there is no constraint to add therefore
% the procedure returns $\ltrue$ at line~\ref{line:mcut-no-v}.
% If $v_r$ does not exist, then it returns constraints encoding
% that if the parent of $v_r$ was not absent, then $v$ must be at a cut at line~\ref{line:mcut-no-vr}.
% If both $v$ and $v_r$ exist, then we add constraints
% that if the parent of $v_r$ is not absent, $v_r$ is
% flagged to be absent in the learned rule
% if and only if $v$ is the cut points at line~\ref{line:mcut-cons}.
% $\textsc{MatchCuts}$ recursively traverses the children of $v$ and $v_r$
% and generates the constraints for the cut pattern.
% The constraints added by the procedure are guarded by the presence of parents
% because we need to enable constraints only when there is a possibility of transition
% from presence to absence.
The call to $\textsc{MatchCuts}$ in \textsc{EncodeP} passes $\lfalse$ as the third input.
Even if the parent of $v_r$ is present,
$v_r$ is a cut point adding cut pattern constraints for the node will create inconsistency.
Therefore, we are passing $\lfalse$.

### 2020-10-15 papers/sugar-synth/variations.tex
variations first cut

In the previous section, we have presented a simplified version of
the synthesis problem.
However, the biological problem has more details.
% There are unsettled modeling choices.
We have developed % and implemented
several variations of the
above method.
% A user may choose the variation depending on the choice of model.

\subsection{Compartments}

The production rules may live in different compartments.
The compartments are ordered.
A molecule moves from one compartment to the next.
The rules of the current compartment apply to the molecule.
We have presented the formal description of the compartments in
section~\ref{sec:model}.
To support the compartments, we take an additional integer
input $k$ in $\textsc{SugarSynth}$ to indicate the maximum number
of compartments.
We construct each template rule with a new variable with domain $[1,k]$
indicating the compartment of the rule.
Let $v_r$ be a node of some template rule. We will write $compart(v_r)$
for the compartment variable for the template rule.

We will alter $\textsc{EncodeProduce}$ and its sub-procedures to
ensure that they enforce the compartment order.
Essentially, we need to encode that a rule is applied when all the
pattern nodes were added in the current or earlier compartment.
For the encoding, we add a new map $compart$ that maps
nodes to variables with domain $[1,k]$.
We modify the function $\textsc{MatchTree}$ by replacing
the {\bf if} by the following code.
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:matchtree-v-absent}
  \If{$isExpand$}
   \State $tCons := ( mark \leq \tau(v)  \land compart(v_r) = compart(v) )$
  \Else
   \State $tCons := ( \tau(v) < mark  \land compart(v_r) \geq compart(v) )$
   \EndIf
   \vspace{1ex}
\end{algorithmic}
In the above code, we have inserted constraints related to compartments
as discussed earlier.
Similarly, we also need to modify the function
$\textsc{EncodeP}$ by inserting the following line after~\ref{line:encodep-ans-match}.
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:encodep-ans-match}
  \State $c \landplus compart(v_r) \geq compart(v')$
  \vspace{1ex}
\end{algorithmic}
Using the above constraints, we can divide the learned rules
into compartments.

@@

\subsection{Fast and slow reactions}
During the stay of a molecule in a compartment, it could be observed that some reactions were dominating than others. This can be explained by characterizing reactions as either slow or fast. A slow reaction can occur only when no other fast reaction is able to extend the molecule in that compartment. We can now define \textsc{Extend} recursively as follows:
\begin{align*}
  \textsc{Extends}(r, m) :=\;&   Apply(m,r) \lor
  \Lor_{\mathclap{i \in [1,w]}} \textsc{Extends}( r, C(m,i ) )\\
    \textsc{Extends}(r, \bot) :=\;&  \nu(v) \neq \bot
  \end{align*}
We now modify $\textsc{EncodeP}$ after~\ref{line:mcut}:
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:mcut}
  \State $c \landplus \lnot \textsc{Fast}(v_r) \implies 
  \Land_{t \in T} (\textsc{Fast}(t) \implies \lnot\textsc{Extends}(t,v))$
  \vspace{1ex}
\end{algorithmic}
Here, \textsc{Fast} sets the constraint on the rule to be fast.
A negative molecule which is a proper subtree of an input molecule will cease to be negative if there is any fast reaction that is able to extend it as fast reactions happen aggressively and can make partial molecules complete. Therefore, we modify the function $\textsc{SugarSynth}$ after~\ref{line:encode-neg-mol}:
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:encode-neg-mol}
  \State $nCons \landplus
  \lnot  \Lor_{r \in R} \textsc{Fast}(r) \land \textsc{Extends}(r,m')$
  \vspace{1ex}
\end{algorithmic}
Here, \textsc{Fast} constraints the rule to be fast. The constraints state that none of the rules should be both fast and able to extend the negative molecule received from the model.

\subsection{Unbounded molecules}
We always observe a finite set of molecules. However,
a set of production rules is capable of producing
an unboundedly large size of molecules. In such cases, rules are allowed to form molecules having repeating patterns of a subtree while rest of the tree being exactly same as one of the input molecules. Constraints are added to encode that those rules are all applied in the same compartment and are slow reactions. We put a limit on the depth of such repeating subtrees ($d_0$) and the number of repetitions ($r_0$) . Then, for a particular depth $d$ and repetitions $r$, we recursively define \textsc{Repeat} and \textsc{ExactMatch} as follows:
\begin{align*}
  \textsc{Repeat}(m, m', r, d) :=\;&  \textsc{Repeat}(m, ancestor(m'), r, d-1)\\
    \textsc{Repeat}(m, m', r, 0) :=\;& (compart(m) = compart(m')) \land \textsc{Repeat}(m, C^{2d}(m'), r-1, d) \\
    \textsc{Repeat}(m, m', 0, d) :=\;&  \textsc{ExactMatch}(m,C^{2d}(m')) \\
    \textsc{Repeat}(\bot, \bot, r, d) :=\;&  \top \\
    \textsc{Repeat}(m, \bot, r, d) :=\;&  \bot \\
    \textsc{Repeat}(\bot, m', r, d) :=\;&  \bot \\
    \textsc{ExactMatch}(m,m') :=\;& (M(m) = M(m')) \land \Land_{i \in [1,w]} \textsc{ExactMatch}(C(m,i),C(m',i)) \\
    \textsc{ExactMatch}(\bot,m') :=\;& \bot \\
    \textsc{ExactMatch}(m,\bot) :=\;& \bot 
  \end{align*}
  where $ancestor(m')$ is the parent node of $m'$ and $C^{2*d}$ is the application of $C(m',i)$, $2d$ times and $i$ comes from the reverse of the path traversed by m' to reach it's $d^{th}$ parent.
The constraints for allowing runaway reactions are added after~\ref{line:consNewR} to modify \textsc{SugarSynth}.
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:consNewR}
  \State $mCons \landplus \Lor_{1 \leq r \leq r_0} \Lor_{1 \leq d \leq d_0} \Lor_{m \in \mu} (M(m) \neq M(\hat{m})) \implies  \textsc{Repeat}(m, \hat{m}, r, d) $
  \vspace{1ex}
\end{algorithmic}
where \textsc{Repeat} adds the constraints described above due to which runaway reactions are allowed.

\subsection{Non-monotonic rules}
Some molecules produced dictate that after a certain rule is applied to them in a compartment, it restricts the application of any other rule from that point of time. Such a rule is said to have \textsc{HardEnds}. This feature of rules help in producing an exact set of desired molecules. We modify the function $\textsc{MatchTree}$ by inserting the following constraints after~\ref{line:matchtree-cons}:
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:matchtree-cons}
  \State $c \landplus \textsc{HardEnds}(v_r) \implies (mark \leq \tau(v))$
  \vspace{1ex}
\end{algorithmic}
In our method, \textsc{HardEnds} sets a constraint on the given rule to have hard ends. The constraints state that if the applicable rule has \textsc{HardEnds}, then it has to be added at a time earlier than the current time of the molecule, effectively restricting the addition of further rules.

>>>>

We have presented a simplified version of
the biological problem.
% However, the biological problem has more details.
We have developed variations of the above method to support more realistic problems.

% \subsection{Compartments}

{\em Compartments : } The production rules may live in different ordered compartments.
% The compartments are ordered.
A molecule moves from one compartment to the next.
The rules of the current compartment apply to the molecule.
% We have presented the formal description of the compartments in
% section~\ref{sec:model}.
To support the compartments, we take an additional integer
input $k$ in $\textsc{SugarSynth}$ to indicate the maximum number
of compartments.
We construct each template rule with a new variable with domain $[1,k]$.
Let $v_r$ be a node of some template rule. We write $compart(v_r)$
for the compartment variable for the template.

We will alter $\textsc{EncodeProduce}$ and its sub-procedures to
ensure that they enforce the compartment order.
We need to encode that a rule is applied when all the
pattern nodes were added in the current or earlier compartment.
% For the encoding, we add a new map $compart$ that maps
% nodes to variables with domain $[1,k]$.
We modify the function $\textsc{MatchTree}$ by replacing
$tCons$ assignment by the following code.\\
\begin{minipage}{1.0\linewidth}
\begin{algorithmic}[1]
  \setcounterref{ALG@line}{line:matchtree-v-absent}
  \State $tCons := isExpand \;?\; ( mark \leq \tau(v)  \land compart(v_r) = compart(v) )$\par
  \mbox{}\qquad\qquad\hspace{10mm} $:( \tau(v) < mark  \land compart(v_r) \geq compart(v) )$
\end{algorithmic}
\end{minipage}
Similarly, we modify the function
$\textsc{EncodeP}$ by inserting the following line after~\ref{line:encodep-ans-match}.\\
\begin{minipage}{1.0\linewidth}
\begin{algorithmic}[1]
  \setcounterref{ALG@line}{line:encodep-ans-match}
  \State $c \landplus compart(v_r) \geq compart(v')$
\end{algorithmic}  
\end{minipage}
% Using the above constraints, we can divide the learned rules
% into compartments.

@@

\paragraph{Fast and slow reactions:}
There is rate associated with chemical reactions. We abstract this by defining slow and fast rules.
The fast rules dominate the slow rule.
A slow rule can occur only when no other fast rule is able to extend the molecule in that compartment.
Let us define function $\textsc{EncodeP}^-$, which generates constraints as $\textsc{EncodeP}$ expect
line~\ref{line:encodep-vr-expand} is missing, i.e., we do not analyze expand part.
% During the stay of a molecule in a compartment, it could be observed that some reactions were dominating than others. This can be explained by characterizing reactions as either slow or fast.
% We can now define \textsc{Extend} recursively as follows:
% $  \textsc{Extends}(r, m) :=\;   Apply(m,r) \lor \Lor_{{i \in [1,w]}} \textsc{Extends}( r, C(m,i ) )$\\
% $\textsc{Extends}(r, \bot) :=\;  \nu(v) \neq \bot$\\
We now modify $\textsc{EncodeP}$ after~\ref{line:encodep-vr-expand} to support fast reactions:\\
\begin{minipage}{1.0\linewidth}
\begin{algorithmic}[1]
  % \vspace{1ex}
  \setcounterref{ALG@line}{line:encodep-vr-expand}
  \State $c \landplus \lnot \textsc{Fast}(v_r) \implies 
  \Land_{t \in T} (\textsc{Fast}(t) \implies \lnot\lor_{ {\ell \in [1,d)}} \textsc{EncodeP}^-( v, t, \ell))$
  % \vspace{1ex}
\end{algorithmic}
  % \vspace{-1mm}
\end{minipage}
Here, \textsc{Fast} sets the constraint on the rule to be fast.
A negative molecule which is a proper subtree of an input molecule will cease to be negative if there is any fast reaction that is able to extend it as fast reactions happen aggressively and can make partial molecules complete. Due to limited space, we will not present the exact constraints.

% Therefore, we modify the function $\textsc{SugarSynth}$ after~\ref{line:encode-neg-mol}:
% \begin{algorithmic}[1]
%   \setcounterref{ALG@line}{line:encode-neg-mol}
%   \State $nCons \landplus
%   \lnot  \Lor_{r \in R} \textsc{Fast}(r) \land \textsc{Extends}(r,m')$
% \end{algorithmic}
% Here, \textsc{Fast} constraints the rule to be fast. The constraints state that none of the rules should be both fast and able to extend the negative molecule received from the model.

{\em Unbounded molecules: }
We only observe a finite set of molecules in cells.
However, a set of production rules may be capable of producing an unboundedly large number of molecules.
In such cases, rules produce molecules the have repeating patterns of a subtree while rest of the tree being exactly same as one of the input molecules. The rules may be acceptable in some biological settings. We modify constraints to not declare such molecules as negative.
Let us suppose we have a repeat pattern of depth $d$ with $r$ repetitions.
Let $v$ be the node in template molecule where repetition has begin.
We define $\textsc{RepeatHeads}(v,r,d)$ that returns nodes $v_0,.... v_r$ such that
$v_1 = v$, $v_i$ is the $d$th ancestor of $v_{i+1}$.
% Those rules should in the same compartment and are slow.
Let us define \textsc{Repeat} and \textsc{Exact}, which encodes that the trees rooted at $v_{i}$s repeat.\\
$\textsc{Repeat}([v_0,...,v_r], v') :=  \textsc{Exact}(v',v_r,\bot) \land \Land_{i\in[0,r-1]} \textsc{Exact}( v_i, v_{i+1},v_{i+1})$\\
% $\textsc{Exact}(v,v') := (M(v) = M(v')) \land \Land_{i \in [1,w]} \textsc{Exact}(C(v,i),C(v',i))$ \\
% $\textsc{Exact}(\bot,v') := \bot, \textsc{Exact}(v,\bot) := \bot, \textsc{Exact}(\bot,\bot) := \bot$\\
$\textsc{Exact}(\bot,v',\_) := \lfalse, \textsc{Exact}(v,\bot,\_) := \lfalse,$\\
$ \textsc{Exact}(\bot,\bot,\_) := \ltrue,\textsc{Exact}(v_s,\_,v_s) := \ltrue$\\
$\textsc{Exact}(v,v',v_s) := (M(v) = M(v')) \land \Land_{i \in [1,w]} \textsc{Exact}(C(v,i),C(v',i),v_s)$\\
% $\textsc{Repeat}(v, v', r, d) :=  \textsc{Repeat}(v, ancestor(v'), r, d-1)$\\
% $\textsc{Repeat}(v, v', r, 0) := (compart(v) = compart(v')) \land \textsc{Repeat}(v, C^{2d}(v'), r-1, d)$ \\
% $\textsc{Repeat}(v, v', 0, d) :=  \textsc{ExactMatch}(v,C^{2d}(v'))$ \\
% $\textsc{Repeat}(\bot, \bot, r, d) :=  \top  \quad \textsc{Repeat}(v, \bot, r, d) :=  \bot
% \quad \textsc{Repeat}(\bot, v', r, d) :=  \bot $\\
% \begin{align*}
%   \textsc{Repeat}(m, m', r, d) :=\;&  \textsc{Repeat}(m, ancestor(m'), r, d-1)\\
%   \textsc{Repeat}(m, m', r, 0) :=\;& (compart(m) = compart(m')) \land \textsc{Repeat}(m, C^{2d}(m'), r-1, d) \\
%   \textsc{Repeat}(m, m', 0, d) :=\;&  \textsc{ExactMatch}(m,C^{2d}(m')) \\
%   \textsc{Repeat}(\bot, \bot, r, d) :=\;&  \top \\
%   \textsc{Repeat}(m, \bot, r, d) :=\;&  \bot \\
%   \textsc{Repeat}(\bot, m', r, d) :=\;&  \bot \\
%   \textsc{ExactMatch}(m,m') :=\;& (M(m) = M(m')) \land \Land_{i \in [1,w]} \textsc{ExactMatch}(C(m,i),C(m',i)) \\
%   \textsc{ExactMatch}(\bot,m') :=\;& \bot \\
%   \textsc{ExactMatch}(m,\bot) :=\;& \bot 
% \end{align*}
  % where $ancestor(m')$ is the parent node of $m'$ and $C^{2*d}$ is the application of $C(m',i)$, $2d$ times and $i$ comes from the reverse of the path traversed by m' to reach it's $d^{th}$ parent.
%   The constraints for allowing runaway reactions are added after~\ref{line:consNewR} to modify \textsc{SugarSynth}.\\
% \begin{algorithmic}[1]
%   \vspace{1ex}
%   \setcounterref{ALG@line}{line:consNewR}
%   \State $mCons \landplus \Lor_{1 \leq r \leq r_0} \Lor_{1 \leq d \leq d_0} \Lor_{m \in \mu} (M(m) \neq M(\hat{m})) \implies  \textsc{Repeat}(m, \hat{m}, r, d) $
%   \vspace{1ex}
% \end{algorithmic}
% where \textsc{Repeat} adds the constraints described above due to which runaway reactions are allowed.
We modify the third constraints of $ \textsc{MolTemplateCorrectness}(\hat{m},\mu)$, which encodes that negative molecules not in $\mu$. We change the definition of $Neq$ as follows.\\
$Neq(v, v') :=\; (\nu(v) \neq M(v') \land \lor_{r\in[1,r_0],d\in[1,d_0]}\textsc{Repeat}(\textsc{RepeatHeads}(v,r,d), v')) $\\
\mbox{}\hspace{30mm}$\lor \Lor_{{i \in [1,w]}} Neq( C(v,i), C(v',i ) ),$\\
where $d_0$ and $r_0$ are limits on the depth of the repeating subtrees and the number of repetitions respectively.
The change will accept molecules with repeating patterns as positive samples.

\paragraph{Non-monotonic rules :}
Some production rules can not be applied if another node is present in a sibling.
We call such rule non-monotonic because it may get disabled as the molecule grows.
This feature of rules help in producing an exact set of desired molecules.
We add an extra bit on each node of rule
template called $\textsc{HardEnds}$.
If the node is absent, its parent is present, and  $\textsc{HardEnds}$ bit is true, then
no node must be present in the matching pattern at the time of the application of the rule.
We modify the function $\textsc{MatchTree}$ by inserting the following constraints after~\ref{line:matchtree-cons}:
\begin{minipage}{1.0\linewidth}
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:matchtree-cons}
  \State $c \landplus \textsc{HardEnds}(v_r) \implies (mark \leq \tau(v))$
\end{algorithmic}  
\end{minipage}
The constraints state that if the applicable rule has \textsc{HardEnds}, then it has to be added at a time earlier than the current time of the molecule, effectively restricting the addition of further rules.
